---
title: "HUNT 1-3-4"
format: html
editor: visual
editor_options: 
  chunk_output_type: console
---

## Information

This script contains the analyses for H1-H3-H4.

## Set environment and load packages

```{r setup}

knitr::opts_chunk$set(
  warning = FALSE,
  message = FALSE
)
```

```{r}

library(dplyr)
library(tibble)
library(lubridate)
library(tidyr)
library(readr)
library(forcats)
library(survival)
```

## Process data

Load data:

```{r}

hunt_1_3_4 <- read_csv("//fil.nice.ntnu.no/nice/p758/data/my_data/hunt_1_3_4.csv")
```

Dependent of H3:

```{r}

hunt_1_3_4_dep <- hunt_1_3_4 |> 
  process_hunt_1_3_4() |> 
  mutate(
    trajectory_group = as_factor(case_when(
        ltpa_group_nt1 == "Low" & ltpa_group_nt3 == "Low" & ltpa_group_nt4 == "Low" ~ "Below recommended, below recommended, below recommended",
        ltpa_group_nt1 == "Low" & ltpa_group_nt3 == "Low" & ltpa_group_nt4 == "Moderate/high" ~ "Below recommended, below recommended, recommended",
        ltpa_group_nt1 == "Low" & ltpa_group_nt3 == "Moderate/high" & ltpa_group_nt4 == "Moderate/high" ~ "Below recommended, recommended, recommended",
        ltpa_group_nt1 == "Low" & ltpa_group_nt3 == "Moderate/high" & ltpa_group_nt4 == "Low" ~ "Below recommended, recommended, below recommended",
        ltpa_group_nt1 == "Moderate/high" & ltpa_group_nt3 == "Low" & ltpa_group_nt4 == "Moderate/high" ~ "Recommended, below recommended, recommended",
        ltpa_group_nt1 == "Moderate/high" & ltpa_group_nt3 == "Moderate/high" & ltpa_group_nt4 == "Moderate/high" ~ "Recommended, recommended, recommended",
        ltpa_group_nt1 == "Moderate/high" & ltpa_group_nt3 == "Moderate/high" & ltpa_group_nt4 == "Low" ~ "Recommended, recommended, below recommended",
        ltpa_group_nt1 == "Moderate/high" & ltpa_group_nt3 == "Low" & ltpa_group_nt4 == "Low" ~ "Recommended, below recommended, below recommended")
    ),
    trajectory_group = relevel(trajectory_group, ref = "Below recommended, below recommended, below recommended")
  )

glimpse(hunt_1_3_4_dep)
```

Independent of H3:

```{r}

# Independent of H3

hunt_1_3_4_indep <- hunt_1_3_4 |> 
  process_hunt_1_3_4() |> 
  mutate(
    trajectory_group = as_factor(case_when(
        ltpa_group_nt1 == "Low" & ltpa_group_nt3 == "Low" & ltpa_group_nt4 == "Low" ~ "Below recommended, below recommended",
        ltpa_group_nt1 == "Low" & ltpa_group_nt3 == "Low" & ltpa_group_nt4 == "Moderate/high" ~ "Below recommended, recommended",
        ltpa_group_nt1 == "Low" & ltpa_group_nt3 == "Moderate/high" & ltpa_group_nt4 == "Moderate/high" ~ "Below recommended, recommended",
        ltpa_group_nt1 == "Low" & ltpa_group_nt3 == "Moderate/high" & ltpa_group_nt4 == "Low" ~  "Below recommended, below recommended",
        ltpa_group_nt1 == "Moderate/high" & ltpa_group_nt3 == "Low" & ltpa_group_nt4 == "Moderate/high" ~ "Recommended, recommended",
        ltpa_group_nt1 == "Moderate/high" & ltpa_group_nt3 == "Moderate/high" & ltpa_group_nt4 == "Moderate/high" ~ "Recommended, recommended",
        ltpa_group_nt1 == "Moderate/high" & ltpa_group_nt3 == "Moderate/high" & ltpa_group_nt4 == "Low" ~ "Recommended, below recommended",
        ltpa_group_nt1 == "Moderate/high" & ltpa_group_nt3 == "Low" & ltpa_group_nt4 == "Low" ~ "Recommended, below recommended")
    ),
    trajectory_group = relevel(trajectory_group, ref = "Below recommended, below recommended")
  )
```

## Analyses

### Un-stratified

#### Crude

##### Dependent of H3

```{r}

hunt_1_3_4_dep_crude_mod <- coxph(Surv(follow_up_time, death_all) ~ trajectory_group + age + sex, id = w22_0389_lopenr_person, data = hunt_1_3_4_dep)

exp(confint(hunt_1_3_4_dep_crude_mod))
```

##### Independent of H3

```{r}

hunt_1_3_4_indep_crude_mod <- coxph(Surv(follow_up_time, death_all) ~ trajectory_group + age + sex, id = w22_0389_lopenr_person, data = hunt_1_3_4_indep)

exp(confint(hunt_1_3_4_indep_crude_mod))
```

#### Multi-adjusted

##### Dependent of H3

```{r}

hunt_1_3_4_dep_multi_mod <- 
  coxph(Surv(follow_up_time, death_all) ~ trajectory_group + age + sex + bp_syst2_nt1blm + bmi_nt1blm + alc_fl2w_nt1blq2 + smo_stat_nt1blq2, id = w22_0389_lopenr_person, data = hunt_1_3_4_dep)

exp(confint(hunt_1_3_4_dep_multi_mod))
```

##### Independent of H3

```{r}

hunt_1_3_4_indep_multi_mod <- 
  coxph(Surv(follow_up_time, death_all) ~ trajectory_group + age + sex + bp_syst2_nt1blm + bmi_nt1blm + alc_fl2w_nt1blq2 + smo_stat_nt1blq2, id = w22_0389_lopenr_person, data = hunt_1_3_4_indep)

exp(confint(hunt_1_3_4_indep_multi_mod))
```

#### Follow-up time and event distribution

##### Dependent of H3

```{r}

# Person-years

calc_person_years(hunt_1_3_4_dep, group_var = "trajectory_group")

# Event distribution

calc_event_dist(hunt_1_3_4_dep, group_var = "trajectory_group") |> 
  filter(death_all == 1)
```

##### Independent of H3

```{r}

# Person-years

calc_person_years(hunt_1_3_4_indep, group_var = "trajectory_group")

# Event distribution

calc_event_dist(hunt_1_3_4_indep, group_var = "trajectory_group") |> 
  filter(death_all == 1)
```

### Stratified

Distinguish the data into two data frames on SES:

```{r}
  
# Dependent of H3

hunt_1_3_4_dep_low <- hunt_1_3_4_dep |> 
  filter(educ_nt1blq2 == "low")

hunt_1_3_4_dep_high <- hunt_1_3_4_dep |> 
  filter(educ_nt1blq2 == "high")
```

```{r}

# Independent of H3

hunt_1_3_4_indep_low <- hunt_1_3_4_indep |> 
  filter(educ_nt1blq2 == "low")

hunt_1_3_4_indep_high <- hunt_1_3_4_indep |> 
  filter(educ_nt1blq2 == "high")
```

#### Multi-adjusted

##### Dependent of H3

Low:

```{r}

hunt_1_3_4_dep_multi_mod_low <- 
  coxph(Surv(follow_up_time, death_all) ~ trajectory_group + age + sex + bp_syst2_nt1blm + bmi_nt1blm + alc_fl2w_nt1blq2 + smo_stat_nt1blq2, id = w22_0389_lopenr_person , data = hunt_1_3_4_dep_low)

exp(confint(hunt_1_3_4_dep_multi_mod_low))
```

High:

```{r}

hunt_1_3_4_dep_multi_mod_high <- 
  coxph(Surv(follow_up_time, death_all) ~ trajectory_group + age + sex + bp_syst2_nt1blm + bmi_nt1blm + alc_fl2w_nt1blq2 + smo_stat_nt1blq2, id = w22_0389_lopenr_person , data = hunt_1_3_4_dep_high)

exp(confint(hunt_1_3_4_dep_multi_mod_high))
```

##### Independent of H3

Low:

```{r}

hunt_1_3_4_indep_multi_mod_low <- 
  coxph(Surv(follow_up_time, death_all) ~ trajectory_group + age + sex + bp_syst2_nt1blm + bmi_nt1blm + alc_fl2w_nt1blq2 + smo_stat_nt1blq2, id = w22_0389_lopenr_person , data = hunt_1_3_4_indep_low)

exp(confint(hunt_1_3_4_indep_multi_mod_low))
```

High:

```{r}

hunt_1_3_4_indep_multi_mod_high <- 
  coxph(Surv(follow_up_time, death_all) ~ trajectory_group + age + sex + bp_syst2_nt1blm + bmi_nt1blm + alc_fl2w_nt1blq2 + smo_stat_nt1blq2, id = w22_0389_lopenr_person , data = hunt_1_3_4_indep_high)

exp(confint(hunt_1_3_4_indep_multi_mod_high))
```

#### Follow-up time and event distribution

##### Dependent of H3

Low:

```{r}

calc_person_years(hunt_1_3_4_dep_low, group_var = "trajectory_group")

calc_event_dist(hunt_1_3_4_dep_low, group_var = "trajectory_group") |> 
  filter(death_all == 1)
```

High:

```{r}

calc_person_years(hunt_1_3_4_dep_high, group_var = "trajectory_group")

calc_event_dist(hunt_1_3_4_dep_high, group_var = "trajectory_group") |> 
  filter(death_all == 1)
```

##### Independent of H3

Low:

```{r}

calc_person_years(hunt_1_3_4_indep_low, group_var = "trajectory_group")

calc_event_dist(hunt_1_3_4_indep_low, group_var = "trajectory_group") |> 
  filter(death_all == 1)
```

High:

```{r}

calc_person_years(hunt_1_3_4_indep_high, group_var = "trajectory_group")

calc_event_dist(hunt_1_3_4_indep_high, group_var = "trajectory_group") |> 
  filter(death_all == 1)
```

## Descriptive statistics

Total number in each SES strata:

```{r}

hunt_1_3_4_dep |> 
  group_by(educ_nt1blq2) |> 
  count()
```

Categorical statistics:

```{r}

# Sex distribution

hunt_1_3_4_dep |> 
  group_by(educ_nt1blq2) |> 
  count(sex)

# Daily smokers

hunt_1_3_4_dep |> 
  group_by(educ_nt1blq2) |> 
  count(smo_stat_nt1blq2)

# Alcohol abstainer

hunt_1_3_4_dep |> 
  group_by(educ_nt1blq2) |> 
  count(alc_fl2w_nt1blq2)

# Consistently below recommended

hunt_1_3_4_dep |> 
  group_by(educ_nt1blq2) |> 
  count(trajectory_group)

# Consistently recommended
  
```

Continuous statistics:

```{r}

hunt_1_3_4_dep |> 
  group_by(educ_nt1blq2) |>
  summarise(
    mean_age = mean(age),
    sd_age = sd(age),
    mean_sbp = mean(bp_syst2_nt1blm),
    sd_sbp = sd(bp_syst2_nt1blm),
    mean_bmi = mean(bmi_nt1blm),
    sd_bmi = sd(bmi_nt1blm)
  )
```

