---
title: "No change in PA w/ maximum follow-up time of 10 years"
format: html
editor: visual
editor_options: 
  chunk_output_type: console
---

## Information

This script contains the analyses for PA as a single baseline measure w/ a maximum follow-up time of 10 years for H1, H3 and H4.

## Set environment and load packages

```{r setup}

knitr::opts_chunk$set(
  warning = FALSE,
  message = FALSE
)
```

```{r}

library(dplyr)
library(tibble)
library(lubridate)
library(readr)
library(tidyr)
library(forcats)
library(survival)
library(purrr)
```

## Load and prepare data

Load data from NTNU server:

```{r}

hunt_1 <- read_parquet("//fil.nice.ntnu.no/nice/p758/data/my_data/hunt_1.parquet")
hunt_3 <- read_parquet("//fil.nice.ntnu.no/nice/p758/data/my_data/hunt_3.parquet")
hunt_4 <- read_parquet("//fil.nice.ntnu.no/nice/p758/data/my_data/hunt_4.parquet")
```

Run data processing functions:

```{r}

hunt_1 <- hunt_1 |>  
  process_hunt_1()

glimpse(hunt_1)

hunt_3 <- hunt_3 |> 
  process_hunt_3()

glimpse(hunt_3)

hunt_4 <- hunt_4 |> 
  process_hunt_4()

glimpse(hunt_4)
```

## Analyses

### Un-stratified

#### H1

##### Crude

```{r}

hunt_1_crude_mod <- coxph(Surv(follow_up_time, death_all) ~ ltpa_hrs_per_week + sex + age, id = w22_0389_lopenr_person, data = hunt_1)

exp(confint(hunt_1_crude_mod))
```

##### Multi-adjusted

```{r}

hunt_1_multi_mod <- coxph(Surv(follow_up_time, death_all) ~ ltpa_hrs_per_week + sex + age + bp_syst2_nt1blm + bmi_nt1blm + alc_fl2w_nt1blq2 + smo_stat_nt1blq2, id = w22_0389_lopenr_person, data = hunt_1)

exp(confint(hunt_1_multi_mod))
```

##### Follow-up time and incidence rate

```{r}


```

#### H3

##### Crude

```{r}

hunt_3_crude_mod <- coxph(Surv(follow_up_time, death_all) ~ ltpa_hrs_per_week + sex + age, id = w22_0389_lopenr_person, data = hunt_3)

exp(confint(hunt_3_crude_mod))
```

##### Multi-adjusted

```{r}

hunt_3_multi_mod <- coxph(Surv(follow_up_time, death_all) ~ ltpa_hrs_per_week + sex + age + bp_syst_mn23_nt3blm + bmi_nt3blm + alc_tot_unit_w_nt3blq1 + smo_stat_nt3blq1, id = w22_0389_lopenr_person, data = hunt_3)

exp(confint(hunt_3_multi_mod))
```

##### Follow-up time and incidence rate

```{r}

calc_person_years(hunt_3)
```

#### H4

##### Crude

```{r}

hunt_4_crude_mod <- coxph(Surv(follow_up_time, death_all) ~ ltpa_hrs_per_week + sex + age, id = w22_0389_lopenr_person, data = hunt_4)

exp(confint(hunt_4_crude_mod))
```

##### Multi-adjusted

```{r}

hunt_4_multi_mod <- coxph(Surv(follow_up_time, death_all) ~ ltpa_hrs_per_week + sex + age + bp_syst_mn23_nt4blm + bmi_nt4blm + alc_tot_unit_w_nt4blq1 + smo_stat_nt4blq1, id = w22_0389_lopenr_person, data = hunt_4)

exp(confint(hunt_4_multi_mod))
```

##### Follow-up time and incidence rate

```{r}

calc_person_years(hunt_4)
```

### Stratified

#### H1

##### Prepare data for the analyses

```{r}

hunt_1_low <- hunt_1 |> 
  filter(educ_nt1blq2 == "low")

hunt_1_high <- hunt_1 |> 
  filter(educ_nt1blq2 == "high")
```

##### Multi-adjusted

Low:

```{r}

hunt_1_multi_mod_low <- coxph(Surv(follow_up_time, death_all) ~ ltpa_hrs_per_week + sex + age + bp_syst2_nt1blm + bmi_nt1blm + alc_fl2w_nt1blq2 + smo_stat_nt1blq2, id = w22_0389_lopenr_person, data = hunt_1_low)

exp(confint(hunt_1_multi_mod_low))
```

High:

```{r}

hunt_1_multi_mod_high <- coxph(Surv(follow_up_time, death_all) ~ ltpa_hrs_per_week + sex + age + bp_syst2_nt1blm + bmi_nt1blm + alc_fl2w_nt1blq2 + smo_stat_nt1blq2, id = w22_0389_lopenr_person, data = hunt_1_high)

exp(confint(hunt_1_multi_mod_high))
```

##### Follow-up time and event distribution

Low:

```{r}

calc_person_years(hunt_1_low)
```

High:

```{r}

calc_person_years(hunt_1_high)
```

#### H3

##### Prepare data for the analyses

```{r}

hunt_3_low <- hunt_3 |> 
  filter(wor_tit_isco1_nt3bli == "low")

hunt_3_high <- hunt_3 |> 
  filter(wor_tit_isco1_nt3bli == "high")
```

##### Multi-adjusted

Low:

```{r}

hunt_3_multi_mod_low <- coxph(Surv(follow_up_time, death_all) ~ ltpa_hrs_per_week + sex + age + bp_syst_mn23_nt3blm + bmi_nt3blm + alc_tot_unit_w_nt3blq1 + smo_stat_nt3blq1, id = w22_0389_lopenr_person, data = hunt_3_low)

exp(confint(hunt_3_multi_mod_low))
```

High:

```{r}

hunt_3_multi_mod_high <- coxph(Surv(follow_up_time, death_all) ~ ltpa_hrs_per_week + sex + age + bp_syst_mn23_nt3blm + bmi_nt3blm + alc_tot_unit_w_nt3blq1 + smo_stat_nt3blq1, id = w22_0389_lopenr_person, data = hunt_3_high)

exp(confint(hunt_3_multi_mod_high))
```

##### Follow-up time and event distribution

Low:

```{r}

calc_person_years(hunt_3_low)
```

High:

```{r}

calc_person_years(hunt_3_high)
```

#### H4

##### Prepare data for the analyses

```{r}

hunt_4_low <- hunt_4 |> 
  filter(educ_nt4blq1 == "low")

hunt_4_high <- hunt_4 |> 
  filter(educ_nt4blq1 == "high")
```

##### Multi-adjusted

Low:

```{r}

hunt_4_multi_mod_low <- coxph(Surv(follow_up_time, death_all) ~ ltpa_hrs_per_week + sex + age + bp_syst_mn23_nt4blm + bmi_nt4blm + alc_tot_unit_w_nt4blq1 + smo_stat_nt4blq1, id = w22_0389_lopenr_person, data = hunt_4_low)

exp(confint(hunt_4_multi_mod_low))
```

High:

```{r}

hunt_4_multi_mod_high <- coxph(Surv(follow_up_time, death_all) ~ ltpa_hrs_per_week + sex + age + bp_syst_mn23_nt4blm + bmi_nt4blm + alc_tot_unit_w_nt4blq1 + smo_stat_nt4blq1, id = w22_0389_lopenr_person, data = hunt_4_high)

exp(confint(hunt_4_multi_mod_high))
```

##### Follow-up time and event distribution

Low:

```{r}

calc_person_years(hunt_4_low)
```

High:

```{r}

calc_person_years(hunt_4_high)
```

### LTPA variable distribution

#### HUNT 1

```{r}

theme_set(theme_minimal())

hunt_1 |> 
  ggplot(aes(x = ltpa_hrs_per_week)) +
  geom_histogram(color = "black", fill = "royalblue") +
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.minor.x = element_blank()
  ) +
  labs(
    title = "Distribution of possible values of LTPA hrs/week",
    subtitle = "HUNT 1",
    x = "LTPA hrs/week",
    y = "Count"
  ) +
  scale_x_continuous(
    breaks = seq(0, 7, 0.5)
  )

ggsave(
  filename = "C:/Users/simen/Desktop/master_thesis/figures/hunt_1_ltpa_dist.jpg",
  dpi = 300,
  height = 8,
  width = 10
)

```

#### HUNT 3

```{r}

theme_set(theme_minimal())

hunt_3 |> 
  ggplot(aes(x = ltpa_hrs_per_week)) +
  geom_histogram(color = "black", fill = "royalblue") +
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.minor.x = element_blank()
  ) +
  labs(
    title = "Distribution of possible values of LTPA hrs/week",
    subtitle = "HUNT 3",
    x = "LTPA hrs/week",
    y = "Count"
  ) +
  scale_x_continuous(
    breaks = seq(0, 7, 0.5)
  )

ggsave(
  filename = "C:/Users/simen/Desktop/master_thesis/figures/hunt_3_ltpa_dist.jpg",
  dpi = 300,
  height = 8,
  width = 10
)
```

#### HUNT 4

```{r}

theme_set(theme_minimal())

hunt_4 |> 
  ggplot(aes(x = ltpa_hrs_per_week)) +
  geom_histogram(color = "black", fill = "royalblue") +
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.minor.x = element_blank()
  ) +
  labs(
    title = "Distribution of possible values of LTPA hrs/week",
    subtitle = "HUNT 4",
    x = "LTPA hrs/week",
    y = "Count"
  ) +
  scale_x_continuous(
    breaks = seq(0, 7, 0.5)
  )

ggsave(
  filename = "C:/Users/simen/Desktop/master_thesis/figures/hunt_4_ltpa_dist.jpg",
  dpi = 300,
  height = 8,
  width = 10
)
```

